En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Supported by organization x.}

### 2019-04-29 mpc19/paper.tex
fixed weak bisim

% Semantics
\newcommand{\semTy}[1]{{\llbracket #1 \rrbracket}_\mathsf{Ty}}
\newcommand{\semMu}[1]{{\llbracket #1 \rrbracket}_\mu}
\newcommand{\semTm}[1]{{\llbracket #1 \rrbracket}_{\lr}}
\newcommand{\semEq}[1]{{\llbracket #1 \rrbracket}_{\LR}}

@@

The coinductive delay datatype was first introduce by Capretta for
representing general recursive functions in Martin-L\"of type theory
\cite{Capretta05}.  Given a type $A$, elements of $\Delay A$ are
possibly non-terminating ``computations'' returning a value of $A$
whenever they terminate. Formally, $\Delay A$ is defined as a
coinductive type with the following introduction rules:
\[
\infer{\now\,a : \Delay A}{a : A}
\quad
\infer={\later\,x : \Delay A}{x : \Delay A}
\]
The constructor $\now$ embeds $A$ into $\Delay A$, so $\now\,a$
represents the terminating computation returning the value $a$. The
constructor $\later$ adds an additional unit of time delay to a
computation. Double rule lines refer to a coinductive constructor,
which can be applied an infinite amount of times.
The non-terminating computation $\never$
is corecursively defined as $\never = \later \,\never$.

@@

Weak bisimilarity is defined in terms of convergence. A computation
$x : \Delay A$ converges to $a : A$ if it terminates in a finite
number of steps returning the value $a$. When this happens, we write
$x \dn a$. The relation $\dn$ is inductively defined by the rules:
\[
\infer{\now\,a \dn a}{}
\quad
\infer{\later\,x \dn a}{x \dn a}
\]
Two computations in $\Delay A$ are weakly bisimilar if they differ by
a finite number of applications of the constructor $\later$. This
informal statement can be formalized in several different but
logically equivalent ways \cite{ChapmanUV19}. 
A possible defintion is as follows:
\begin{equation}\label{eq:weak1}
x \approx y = (a : A) \to x \dn a \leftrightarrow y \dn a
\end{equation}
The definition says that $x$ and $y$ are weakly bisimilar if, whenever
$x$ terminates returning a value $a$, then $y$ also terminates
returning $a$, and vice versa.
Alternatively, weak bisimilarity can be coinductively formulated as follows:
\begin{equation}
\label{eq:weak2}
\arraycolsep=8pt\def\arraystretch{2.8}
\begin{array}{cc}
\infer{\now_{\approx}  : \now\,a \approx \now\,a}{}
&
\infer={\later_{\approx} \,p : \later\,x_1 \approx \later\,x_2}{p : x_1 \approx x_2}
\\
\infer{\laterL_{\approx} \, p: \later\,x \approx \now\,a}{p : x \approx \now \,a}
&
\infer{\laterR_{\approx} \, p: \now\,a \approx \later\,x}{p : \now\,a \approx x}
\end{array}
\end{equation}
Notice that the constructor $\later_{\approx}$ is
coinductive. Formulation (\ref{eq:weak1}) and (\ref{eq:weak2}) are
logically equivalent, meaning that two computations are weakly
bisimilar as in (\ref{eq:weak1}) if and only if they are weakly
bisimilar as in (\ref{eq:weak2}).
In the rest of the paper we will employ
both encodings of weak bisimilarity, in each case explicitly stating
if we are working with definition (\ref{eq:weak1}) or definition (\ref{eq:weak2}).
Weak bisimilarity is an equivalence
relation and it is a congruence w.r.t. the $\later$ operation.
For example, here is a proof that weak bisimilarity is reflexive,
employing definition (\ref{eq:weak2}).
\begin{align*}
& \refl_{\approx} : \{x : \Delay A\} \to x \approx x \\
& \refl_{\approx} \,\{\now\,a\} = \now_\approx \\
& \refl_{\approx} \,\{\later\,x\} = \later_\approx\,(\refl_\approx \,\{x\})
\end{align*}

@@

As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
up to weak bisimilarity. Here we show the proof of the unfolding
axiom. We employ the definition (\ref{eq:weak2}) of weak
bisimilarity. The proof relies on an auxiliary proof $\unfolding'$. 
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (B + A)) \to
  \bind\, \copairD {\now}{\iterD \,f} \approx \iterD' \,f} \\
\unfolding' \,f \,(\now \,(\inl \,b)) & = \now_\approx \\
\unfolding' \,f \,(\now \,(\inr \,a)) & = \laterR_{\approx} \,
  \refl_\approx \\
\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
  x) \\
 \\
\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (B + A)) \to
  \copairD {\now}{\iterD \,f}\diamond f \approx \iterD \,f}\\
\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
\end{array}
\]

@@

@@

Summing up, in this section we have showed that the $\traceD$ operator
can be restricted to act on partial isomorphisms. That is, the
following type is inhabited:
\[
\mathsf{semTrace} : A + C\simeq B + C \to A \simeq C
\]

@@

Terms of \Pio\ are modelled as morphism of $\Inv\Dapprox$,
i.e. partial isomorphisms. Formally, we define an operation:
\[
\semTm - : (A \lr B) \to \semTy A \, \rho \simeq \semTy B \, \rho
\]
We refer the interested reader to our Agda formalization for a
complete definition of the interpretation of terms. We only emphasize
the semantics of the trace operator: $\semTm {\trace\,f} =
\mathsf{semTrace}\,\semTm f$.

>>>>

% Semantics
\newcommand{\semTy}[1]{{\llbracket #1 \rrbracket}_\mathsf{Ty}}
\newcommand{\semMu}[1]{{\llbracket #1 \rrbracket}_\mu}
\newcommand{\semTm}[1]{{\llbracket #1 \rrbracket}_{\lr}}
\newcommand{\semEq}[1]{{\llbracket #1 \rrbracket}_{\LR}}
\newcommand{\semTrace}{\mathsf{trace}_{\D\simeq}}

@@

The coinductive delay datatype was first introduce by Capretta for
representing general recursive functions in Martin-L\"of type theory
\cite{Capretta05}.  Given a type $A$, elements of $\Delay A$ are
possibly non-terminating ``computations'' returning a value of $A$
whenever they terminate. Formally, $\Delay A$ is defined as a
coinductive type with the following introduction rules:
\[
\infer{\now\,a : \Delay A}{a : A}
\quad
\infer={\later\,x : \Delay A}{x : \Delay A}
\]
The constructor $\now$ embeds $A$ into $\Delay A$, so $\now\,a$
represents the terminating computation returning the value $a$. The
constructor $\later$ adds an additional unit of time delay to a
computation. Double rule lines refer to a coinductive constructor,
which can be employed an infinite number of times in the construction
of a term of type $\Delay A$.  E.g., the non-terminating computation
$\never$ is corecursively defined as $\never = \later \,\never$.

@@

Weak bisimilarity is defined in terms of convergence. A computation
$x : \Delay A$ converges to $a : A$ if it terminates in a finite
number of steps returning the value $a$. When this happens, we write
$x \dn a$. The relation $\dn$ is inductively defined by the rules:
\[
\infer{\now\,a \dn a}{}
\quad
\infer{\later\,x \dn a}{x \dn a}
\]
Two computations in $\Delay A$ are weakly bisimilar if they differ by
a finite number of applications of the constructor
$\later$. Alternatively, we can say that two computations $x$ and $y$
are weakly bisimilar if, whenever $x$ terminates returning a value
$a$, then $y$ also terminates returning $a$, and vice versa. This
informal statement can be formalized in several different but
logically equivalent ways \cite{ChapmanUV19}. Here we consider a
coinductive formulation employed e.g. in \cite{Danielsson18}.
%% A possible defintion is as follows:
%% \begin{equation}\label{eq:weak1}
%% x \approx y = (a : A) \to x \dn a \leftrightarrow y \dn a
%% \end{equation}
%% The definition says that $x$ and $y$ are weakly bisimilar if, whenever
%% $x$ terminates returning a value $a$, then $y$ also terminates
%% returning $a$, and vice versa.
%% Alternatively, weak bisimilarity can be coinductively formulated as follows:
\begin{equation}
\label{eq:weak2}
\arraycolsep=8pt\def\arraystretch{2.8}
\begin{array}{cc}
\infer{\now_{\approx}  : \now\,a \approx \now\,a}{}
&
\infer={\later_{\approx} \,p : \later\,x_1 \approx \later\,x_2}{p : x_1 \approx x_2}
\\
\infer{\laterL_{\approx} \, p: \later\,x \approx \now\,a}{p : x \approx \now \,a}
&
\infer{\laterR_{\approx} \, p: \now\,a \approx \later\,x}{p : \now\,a \approx x}
\end{array}
\end{equation}
Notice that the constructor $\later_{\approx}$ is coinductive. This
allows us to prove $\never \approx \never$. 
%% Formulation
%% (\ref{eq:weak1}) and (\ref{eq:weak2}) are logically equivalent,
%% meaning that two computations are weakly bisimilar as in
%% (\ref{eq:weak1}) if and only if they are weakly bisimilar as in
%% (\ref{eq:weak2}).  In the rest of the paper we will employ both
%% encodings of weak bisimilarity, in each case explicitly stating if we
%% are working with definition (\ref{eq:weak1}) or definition
%% (\ref{eq:weak2}). 
 Weak bisimilarity is an equivalence relation and it
is a congruence w.r.t. the $\later$ operation.  For example, here is a
proof that weak bisimilarity is reflexive.
\begin{align*}
& \refl_{\approx} : \{x : \Delay A\} \to x \approx x \\
& \refl_{\approx} \,\{\now\,a\} = \now_\approx \\
& \refl_{\approx} \,\{\later\,x\} = \later_\approx\,(\refl_\approx \,\{x\})
\end{align*}

@@

As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
up to weak bisimilarity. Here we show the proof of the unfolding
axiom. The proof relies on an auxiliary proof $\unfolding'$. 
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (B + A)) \to
  \bind\, \copairD {\now}{\iterD \,f} \approx \iterD' \,f} \\
\unfolding' \,f \,(\now \,(\inl \,b)) & = \now_\approx \\
\unfolding' \,f \,(\now \,(\inr \,a)) & = \laterR_{\approx} \,
  \refl_\approx \\
\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
  x) \\
 \\
\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (B + A)) \to
  \copairD {\now}{\iterD \,f}\diamond f \approx \iterD \,f}\\
\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
\end{array}
\]

@@

We conclude this section by remarking that the construction of a trace
operator in the Kleisli category of the maybe monad is impossible
without the assumption of additional classical principles. In fact,
given a map $f : A + C \to B + C + 1$, let $xs$ be the possibly
infinite sequence of elements of $B + C + 1$ produced by the iteration
of $f$ on a given input $A + C$. In order to construct the trace of
$f$, we need to decide whether $xs$ is a finite sequence terminating
with an element of $B + 1$, or $xs$ is an infinite stream of elements
of $C$. This decision requires the limited principle of omniscience,
an instance of the law of excluded middle not provable in Martin-L\"of
type theory.
\[
\mathsf{LPO} = (s : \mathbb{N} \to 2) \to ((n : \mathbb{N}) \times
s\,n \equiv \mathsf{true}) + ((n : \mathbb{N}) \to s\,n \equiv \mathsf{false})
\]

@@

Summing up, in this section we have showed that the $\traceD$ operator
can be restricted to act on partial isomorphisms. That is, the
following type is inhabited:
\[
\semTrace : A + C\simeq B + C \to A \simeq C
\]

@@

Terms of \Pio\ are modelled as morphism of $\Inv\Dapprox$,
i.e. partial isomorphisms. Formally, we define an operation:
\[
\semTm - : (A \lr B) \to \semTy A \, \rho \simeq \semTy B \, \rho
\]
We refer the interested reader to our Agda formalization for a
complete definition of the interpretation of terms. Here we limit
ourselves to emphasize
the semantics of the trace operator: $\semTm {\trace\,f} =
\semTrace\,\semTm f$.
